---
title: Lab 1
author: Ansel George
output: pdf_document
---

```{r, message=F}
library(bipartite)
library(igraph)

set.seed(10)
```


# Problem 4

*What can edge weights be in ecological networks? Try to think of at least 3
different measures for an ecological interaction. If you work with other
network types, answer the question according to your domain of expertise.*

1. Migration rates between demes.
2. Transition probabilities in a Markov chain (generalization of 1.)
3. Energy/mass flow in a food web.


# Problem 6

*What are the advantages/disadvantages of edge lists compared to matrices?*

Edge lists are more memory-efficient for sparse matrices. If the number of
edges $E$ is small enough that $E << V$, where $V$ is the number of vertices,
then. For dense matrices, though, the number of edges could be $\sim N^2$, and
the advantage is lost.


# Problem 7

*Try programming a function that takes a matrix and returns an edge list. Note
that it may be important if the matrix is directed or weighted.*

```{r}
returnEdgeList <- function(M, directed=T) {
  if (directed==F) {
    E <- which(M != 0, arr.ind=T)
    E <- E[E[,2] >= E[,1],] # record edges in upper triangle (assume symmetry)
    weight <- M[E]
    E <- cbind(E, weight)
    colnames(E) <- c("from", "to", "weight")
  } else {
    E <- which(M != 0, arr.ind=T)
    weight <- M[E]
    E <- cbind(E, weight)
    colnames(E) <- c("from", "to", "weight")
  }
  return(E)
}
```

```{r}
M <- matrix(nrow=4, ncol=4, byrow=T,
            c(1, 3, 0, 0,
              4, 0, 0, 1,
              0, 0, 3, 0,
              0, 0, 0, 1))

E <- returnEdgeList(M)
E
```

# Problem 8

*Try programming a function that takes an edge list and returns a matrix. Note
that it may be important if the matrix is directed or weighted.*

```{r}
returnAdjacencyMatrix <- function(edgelist, directed=T) {
  maxdim <- max(edgelist[,1:2])
  M <- matrix(nrow=maxdim, ncol=maxdim, rep(0, maxdim*maxdim))
  if (directed == T) {
    if (dim(edgelist)[2] == 3) {
       M[edgelist[,1:2]] <- edgelist[,3]
    } else if (dim(edgelist)[2] == 2) {
       M[edgelist[,1:2]] <- 1
    }
  } else {
    # preserve symmetry for directed graph.
    if (dim(edgelist)[2] == 2) {
      M[edgelist[,1:2]] <- 1
      M[edgelist[,2:1]] <- 1
    } else if (dim(edgelist)[2] == 3) {
      edgelist <- edgelist[edgelist[,2] >= edgelist[,1],] # record edges in upper triangle (assume symmetry)
      M[edgelist[,1:2]] <- edgelist[,3] # assume weights are correct
      M[edgelist[,2:1]] <- edgelist[,3]
    }
  }
  return(M)
}
```

```{r}
returnAdjacencyMatrix(E)
```


# Problem 9
*Now try to load a different data set. Where can you find available ones?*

The `bipartite` packages has several other datasets available. A list is
available on the packages documentation:

\url{https://cran.r-project.org/web/packages/bipartite/bipartite.pdf}

For example, the `barrett1987` dataset contains information about pollinators
in the boreal forest in New Brunswick. 

```{r}
data(barrett1987)
```

Other datasets are available online (datadryad, Interaction Web DataBase, and so on).


# Problem 17

*Try to obtain 2 data sets from 2 different repositories, load them and plot
them.*


## Species-specific responses to habitat conversion

Frishkoff L, Karp D (2019) Data from: Species-specific responses to habitat
conversion across scales synergistically restructure Neotropical bird
communities. Dryad Digital Repository.
\url{https://doi.org/10.5061/dryad.fg8kq66}


Download data:

```{bash, message=F}
mkdir -p frishkoff-data
cd frishkoff-data
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212360/community-data.csv
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212361/edge-data.csv
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212362/forestcover-data.csv
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212363/precip-data.csv
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212368/visit-data.csv
wget -nc https://datadryad.org/bitstream/handle/10255/dryad.212369/readme.txt
cd ..
```

Load data:

```{r}
prefix <- "frishkoff-data/"
community_data <- read.csv(paste(prefix, "community-data.csv", sep=""))
edge_data <- read.csv(paste(prefix, "edge-data.csv", sep=""))
forestcover_data <- read.csv(paste(prefix, "forestcover-data.csv", sep=""))
precip_data <- read.csv(paste(prefix, "precip-data.csv", sep=""))
visit_data <- read.csv(paste(prefix, "visit-data.csv", sep=""))
```

Plot data:

```{r}
sites <- community_data$Site
species <- community_data$Species
links <- data.frame(species, sites)

y <- c(rep(0, length(levels(species))), rep(1, length(levels(sites))))
# x <- runif(length(y))
x <- c(seq(1, length(levels(species)))/length(levels(species)),
       seq(1, length(levels(sites)))/length(levels(sites)))
coords <- cbind(x, y)
nodes <- c(as.vector(unique(species)), as.vector(unique(sites)))

p <- graph.data.frame(links, vertices=nodes, directed=F)
par(mar = c(0, 0, 0, 0)) # reduce margin size
# plot(p, layout=coords, vertex.color=(coords[,2]+1), vertex.label=NA, vertex.size=5, edge.width=.1)
plot(p, layout=coords, vertex.color=(coords[,2]+1), vertex.label=NA,
     vertex.size=5, edge.width=.1,
     edge.color = adjustcolor("black", alpha.f = .025))
```

The top set (blue) are geographical sites, and the bottom (orange) are species.
An edge between site and species signifies that the species is recorded to have
been found in the site. There are 25 sites and 150 species. One can vaguely
tell by the graph that some species are found in more sites and some site are
host to more species. Surely there are better ways to represent this clearly...


## Rainforest ant-plant interaction network

This describes de ant-plant interaction network in the rainforest at the
Australian Canopy Crane in Cape Tribulation, Far North Queensland, Australia
and adjacent forests within 5 km radius of the crane site, including patches of
open secondary forest. Counts of ant colonies attending extrafloral (EFN) and
floral nectaries (FN) are provided, as well as independent abundance estimates
for ant colonies based on sugar bait experiment (from Blüthgen and Fiedler
2004). Ants with zero abundance are those not recorded during bait experiments.
For plants, number of plant individuals on which any insect were recorded on
nectaries given as abundance estimate.  Plant life forms: cl = climber, he =
herb, sh = shrub, tr = tree, pa = palm. 

```{bash, message=F}
mkdir -p bluthgen-data
cd bluthgen-data
wget -nc https://www.nceas.ucsb.edu/interactionweb/data/ant_plant/text_matrices/bluthgen_et_al_2004.txt
wget -nc https://www.nceas.ucsb.edu/interactionweb/data/ant_plant/excel_matrices/bluthgen_2004.xls
cd ..
```

```{r}
path <- "bluthgen-data/"
filename <- "bluthgen_et_al_2004.txt"
interaction_data <- read.table(paste(path, filename, sep=""), sep = '\t', header = F)
ant_data <- read.table("bluthgen-data/ants.txt", sep = '\t', header = T, quote = "",)
plant_data <- read.table("bluthgen-data/plants.txt", sep = '\t', header = T)
```

```{r, message=F, warning=F}
colnames(interaction_data) <- plant_data$Species
rownames(interaction_data) <- ant_data$Species
visweb(interaction_data)
visweb(interaction_data > 0) # binary version

plotweb(interaction_data)
```
