---
title: Lab 2
author: Ansel George
output: pdf_document
---

```{r, message=F}
library(igraph)
library(bipartite)
library(ggplot2)

set.seed(10)
```


# Problem 4
*Try to calculate and access metrics at the species level (e.g., host assemblage
or parasite assemblage) level. Use the help to discover which metrics bipartite
has implemented.*

For host vs parasite assemblage statistics, use the `grouplevel` function from
`bipartite`.

```{r}
data(olesen2002flores)
olesen2002flores_binary <- 1*(olesen2002flores>0)

gl_metrics <- grouplevel(olesen2002flores)
gl_metrics
```

The first column corresponds to the parasite statistics and the second to those
for the hosts. Several of these parameters can be computed from the compute
cross products on the binary data. For the host assemblage, the entries reveal
the number of shared parasites a pair of hosts have, and for the parasite
assemblage, the entries represent the number of hosts a pair of parasite both
parasitize.


```{r}
# Host assemblage
olesen2002flores_hosts <- tcrossprod(olesen2002flores_binary)
weighted.mean(colSums(olesen2002flores_binary),
              w=apply(olesen2002flores,2,sum)/sum(olesen2002flores)) # mean.number.of.links.HL
mean(rowSums((olesen2002flores_hosts-diag(10)*diag(olesen2002flores_hosts))))/9 # mean.number.of.shared.partners.LL

# Parasite assemblage
olesen2002flores_parasite <- crossprod(olesen2002flores_binary)
weighted.mean(rowSums(olesen2002flores_binary),
              w=apply(olesen2002flores,1,sum)/sum(olesen2002flores)) # mean.number.of.links.HL
mean(rowSums((olesen2002flores_parasite-diag(12)*diag(olesen2002flores_parasite))))/11 # mean.number.of.shared.partners.HL
```


# Problem 7
*What information do we lose when projecting?*

All the information regarding specific interactions between higher-lower
species is lost.


# Problem 9
*What is the ecological interpretation of the projections of a plant-pollinator
network and of a host-parasite network?*

**Plant-pollinator:**

1. Plant projection - number of shared pollinators between species.
2. Pollinator projection - number of plants a pair of species both pollinate.

**Host-parasite:**

1. Host projection - number of shared parasites for a given pair of hosts.
2. Parasite projection - number of hosts a pair of parasites both exploit.


# Problem 14
*Can you think of an ecological interpretation to each of these measures? Some
intuition is given in the Martín-González et al. (2010) and Delmas et al.
(2019) references.*

Closeness centrality measures the proximity of a species to others. If one
consider flows of information through the network where signals are split or
attenuated as they hop from one node to the next, then species with high
connectedness centrality will receive much more information then less connected
species. The information could represent perturbations in the food chain, such
as extinctions, population explosions, etc.

Betweenness centrality measures the importance of a species in terms of some
metric for flow through the network. One metric can be the minimum number of
links to traverse the food web from any given pair of species. The measure is
not unrelated from connectedness centrality. For example, betweenness and
connectedness can identify keystone species given a network topology where
nodes are sparsely connected absent those with high connectedness/betweenness.
On the other hand, the two can differ substantially in very dense networks
where many nodes are connected to many others. All will have high connectedness
centrality but potentially low betweenness centrality.

Eigenvector centrality is a way of interpreting flows through the food web. In
a network, eigenvectors represent flow proportions per node that maximize
overall flow in the network. If this is taken as a proxy for
importance/centrality, then the values of the eigenvector correspond to the
centrality of a given species. Interpreting the centrality measure (aka
eigenvalue) ecologically is difficult and depends on network structure. If the
distribution of normalized loadings for each node is skewed in favor of a small
proportion of them, the these nodes are likely to be highly important to the
overall network, and if they are all roughly equal, there are many connections
and redundant paths through the network.

Note that this requires the network edges be non-negative the network itself
irreducible. If the network allows negative interactions/flows or if orthogonal
subgraphs in the network exist, then this approach may not necessarily hold.


# Problem 17
*What is the dominant sub-graph in the Otago food web? What is its ecological
meaning?*

There are two motifs in the subgraph with proportion greater than 0.1: 3 and 7.

Motif 7 (0.4170122917): One plant is pollinated by two different pollinators.

Motif 3 (0.2357665334): Two plants are pollinated by the same pollinator.
