---
title: Lab 4
author: Ansel George
output: pdf_document
---

```{r, message=F}
library(bipartite) # will also load vegan
library(ggplot2)
library(knitr)

set.seed(10)
```

# Problem 1
*What kind of model (equiprobable, probabilistic, fixed) is `r00`?*

equiprobable (row and column)


# Problem 2
*Can you program the algorithm for `r00` by yourself?*

```{r}
r00 <- function(A) {
  nr <- dim(A)[1]
  nc <- dim(A)[2]
  out <- rep(0, nr*nc)
  out[sample.int(nr*nc, sum(A))] <- 1
  dim(out) <- c(nr, nc)
  return(out)
}
```

```{r}
nr <- 5
nc <- 4
A <- matrix(sample(c(0,1), nc*nr, replace=T), nrow=nr, ncol=nc)

A
r00(A)
```


# Problem 4
*There is no better way to understand the mechanics of randomization algorithms
than to program one for yourself! Try to program the probabilistic model from
[6]: "The probability of each cell being occupied is the average of the
probabilities of occupancy of its row and column." This means that the
probability of drawing an interaction is proportional to the degree of both the
lower-level and the higher-level species. Mathematically:*

\begin{align}
P_{ij} = \frac{1}{2}\left(\frac{\sum A_{i}}{J}+\frac{\sum A_{j}}{I}\right)
\end{align}

_where $A_{ij}$ is the incidence matrix in which $I$ and $J$ denote the total
number of species in the rows and columns, respectively. Try to program that
algorithm. Can you note one problem that you are facing?_

```{r}
randomize <- function(A) {
  I <- dim(A)[1]
  J <- dim(A)[2]
  a1 <- rowSums(A) / J
  a2 <- colSums(A) / I
  out <- 1/2 * (outer(a1, rep(1,length(a2))) +
                t(outer(a2, rep(1, length(a1)))))
  out <- matrix(rbinom(J*I, 1, as.vector(out)), nrow=I, ncol=J)
  return(out)
}

randomize2 <- function(A) {
  I <- dim(A)[1]
  J <- dim(A)[2]
  a1 <- rowSums(A) / J
  a2 <- colSums(A) / I
  p <- 1/2 * (outer(a1, rep(1,length(a2))) +
              t(outer(a2, rep(1, length(a1)))))
  p <- as.vector(p) / sum(p)
  out <- rep(0, nr*nc)
  out[sample.int(nr*nc, sum(A), replace=F, p)] <- 1
  dim(out) <- c(nr, nc)
  return(out)
}
```

```{r}
A
A_r <- randomize(A)
A_r
```

```{r}
tmp <- matrix(0, nrow=nr, ncol=nc)
n <- 10000
for (i in 1:n) {
  tmp <- tmp + randomize(A)
}
tmp <- tmp / n
```


The algorithm will not necessarily preserve the marginal likelihoods. Their
expectations will match the values of the original matrix, but there will be
deviations from the expectation arising from the probabilistic way the
connections are generated.

```{r}
tmp <- matrix(0, nrow=nr, ncol=nc)
n <- 25000
for (i in 1:n) {
  tmp <- tmp + randomize(A)
}
tmp <- tmp / n
```

```{r}
I <- dim(A)[1]
J <- dim(A)[2]
a1 <- rowSums(A) / J
a2 <- colSums(A) / I
p <- 1/2 * (outer(a1, rep(1,length(a2))) +
            t(outer(a2, rep(1, length(a1)))))

# Compare tmp and p
tmp
p
```


The algorithm could be alternatively formulated as follows:

```{r}
randomize2 <- function(A) {
  I <- dim(A)[1]
  J <- dim(A)[2]
  a1 <- rowSums(A) / J
  a2 <- colSums(A) / I
  p <- 1/2 * (outer(a1, rep(1,length(a2))) +
                t(outer(a2, rep(1, length(a1)))))
  p <- as.vector(p) / sum(p)
  out <- rep(0, nr*nc)
  out[sample.int(nr*nc, sum(A), replace=F, p)] <- 1
  dim(out) <- c(nr, nc)
  return(out)
}
```

This version will preserve the number of interactions, but it will skew the
probability of an interaction for each element of A because of the
combinatorial problem of drawing from a set of values without replacement.

```{r}
tmp <- matrix(0, nrow=nr, ncol=nc)
n <- 25000
for (i in 1:n) {
  tmp <- tmp + randomize2(A)
}
tmp <- tmp / n
```

```{r}
I <- dim(A)[1]
J <- dim(A)[2]
a1 <- rowSums(A) / J
a2 <- colSums(A) / I
p <- 1/2 * (outer(a1, rep(1,length(a2))) +
            t(outer(a2, rep(1, length(a1)))))

# Compare tmp and p
tmp
p
```

# Problem 5
*Try comparing to a fixed shuffling algorithm like `curveball`. Are there
differences in the results when using different null models? Why?*

The `curveball` fixed shuffling algorithm will preserve the number of links and
the marginal sums. Other algorithms, such as `r1` and `r0`, will instead fix
the marginals. Different algorithms will impose different constraints on what
values individual elements can take given other elements, so they will produce
different randomized results. See below:

```{r}
num_iterations <- 25000

#
# r00
#

null_r00 <- vegan::nullmodel(A, method = 'r00') # Select a null model for the data
shuffled_r00 <- simulate(null_r00, nsim = num_iterations) # Shuffle
shuffled_r00[,,1]

# r00 has equiprobable elements (sum(A) / (nc*nr))
apply(shuffled_r00, c(1,2), mean)
sum(apply(shuffled_r00, c(1,2), mean))


#
# curveball
#

null_curveball <- vegan::nullmodel(A, method = 'curveball') # Select a null model for the data
shuffled_curveball <- simulate(null_curveball, nsim = num_iterations) # Shuffle
shuffled_curveball[,,1]
apply(shuffled_curveball, c(1,2), mean)
sum(apply(shuffled_curveball, c(1,2), mean))
```


```{r, message=F, results="hide"}
purl("lab4.Rmd", output = "lab4.R")
```
